Maximum Entropy Spectral Analysis: a case study

### 2021-04-15 paper/mesa.tex
First bunch of edits

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ABSTRACT
\begin{abstract}
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package `memspectrum`. We characterize our implementation by performing a power spectral density analysis on syntethic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using both syntethic data and real gravitational wave data from the LIGO-Virgo collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime.
%Several methods are characterized on synthetic noise having a known power spectral density (PSD) via the computation of the percentage error and then compared.

@@

The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, astronomy as well as engineering. Of particular interest for several areas of physics is the study of stochastic time series. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\cite{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will explore further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

@@

The choice of a window function is arbitrary and typically is made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of PSD is reached. A high frequency resolution implies high variance and vice-versa.
Besides the window function, Welch' method requires a number of arbitrary choices to be made, since the number of time slices and the overlap between consecutive slices. All the knobs must be tuned by hand and their choice can dramatically affect the PSD estimation.
The final PSD estimate depends on all such choices, hence begging the question of what the ``real" PSD is.

@@

In this paper, we discuss the details of the Maximum entropy principle, its application to the problem of PSD estimation with Burg algorithm and the link between Burg's algorithm and autoregressive process.
Our goal is to bring (again) to public attention Maximum Entropy Spectral analysis, in the hope that it will be widely employed as a way out to the many problem posed by Welch's algorithm (or other similar methods).
To facilitate this goal, we present and describe a new code, \texttt{memspectrum}, that provides a robust and easy-to-use python implementation of the algorithm\footnote{
Available at link: \url{https://pypi.org/project/memspectrum/}.
}.
We provide a thorough assessment of the performance of our code and we validate our results performing a number of tests on simulated and real data.
We also compare our results with those of spectral analysis carried out with the standard Welch method.
In order to apply our model on a realistic setting, we analyse some time series of broad interest in the scientific community.

Our paper is organized as follows: we being by briefly reviewing the theoretical foundations of the maximum entropy principle in Sec.~\ref{sec:foundations}. Sec.~\ref{sec:validation} presents the validation of Burg's method as well as of our implementation on simulated data. In Sec.~\ref\label{sec:Welch_comparison} we compare the results from \texttt{memspectrum} with the Welch method; Sec.~\ref{sec:applications} presents a few applications to time series and, finally, we conclude with a discussion in Sec.~\ref{sec:conclusion}.

@@

It is very common that the constraints provided by the evidence are not enough for setting the probabilities for each event: in this case, it is reasonable to assume that the probability assignment should make the experiment as unpredictable as possible. In other words, the amount of `information' introduced by the the probability assignment should be as high as possible.

%Let us now reverse the perspective: what is the most unbiased probability assignment we can make if all we know is that one over $N$ outcomes will occur? Since we have no reason to prefere one outcome over the others, the least biased assignment is the one equivalent to (\ref{eq:indifference}) and is found to be coincident with the most uncertain experiment. This result represents the core idea of MAXENT principle: the least biased probability assignment can be obtained maximizing the "expected" information's entropy. \\ 
%In the previous example we had no reason to prefere one outcome over the others but that's not always the case. This hypothesis can be relaxed introducing the concept of 'evidence', that represents the knowledge we posses on the system under study. With proper evidence some outcomes can be preferred to others. Probability assignment must obey to all the evidence we possess.

MAXENT puts the aforementioned principle into a more precise mathematical context by stating that probabilities should be assigned by maximizing uncertainty (information entropy) using evidence as constraint. 
This defines a variational problem, where a numerical measure of uncertainty $H\left[p_1, \dots, p_N\right]$ has to maximized.

@@

Of particular importance is the case in which, in addition to the mean, also the variance is known. MAXENT leads to the Gaussian distribution. 
We retain this particularly interesting from the foundational point of view, since it provides a deeper insight into the ubiquitous Gaussian distribution.
It is not only the limit distribution provided by the central limit theorem for finite variance processes but it is also the distribution that maximizes the entropy. For this reason, appealing to MAXENT principle, it is the correct assignment if mean and covariance are the only quantities that fully define our process. In some sense, we can interpret the central limit theorem as the natural `statistical' evolution toward a configuration that maximizes entropy.

For this work, we are particularly interested in the mutli-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realisation of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of auto-covariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution\cite{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

We have already seen that for a wide-sense stationary process the mean function is independent of time, hence redefined to be equal to zero without loss of generality, and the auto-covariance function is dependent only on the time lag $\tau \equiv t_i - t_j$. One can thus choose a sampling rate $\Delta t$ so that $C_{ij} = C((i-j)\Delta t)$. The autocovariance matrix thus becomes a Toeplitz matrix\footnote{
We remind the reader that a Toeplix matrix is a matrix in the form:
\begin{pmatrix}
	a_0 & a_1 & a_2 & \ldots & \ldots& \ldots &a_n\\
	a_{-1} & a_0 & a_1 & \ldots & \ldots& \ldots &a_{n-1}\\
	a_{-2} & a_{-1} & a_0 & \ldots & \ldots & \ldots  &a_{n-2}\\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots &\vdots\\
	a_{-n +1} & \ldots & \ldots & \ldots& a_{-1} & a_0    &a_{1}\\
	a_{-n} & \ldots & \ldots & \ldots& a_{-2} & a_{-1} & a_0
\end{pmatrix}

}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalised by the discrete Fourier transform base\cite{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} {x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\cite{prob_information_theory}
%P. M. Woodward, D. W. Fry, and W. Higinbotham, Probability and information theory, with applications to radar; 2nd ed. (Pergamon Press, Oxford, 1964), URL http://cds.cern.ch/record/2031792.
and of gravitational waves data analysis, e.g. \cite{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the auto-covariance function.

@@

As in the previous examples, one needs to set up a variational problem where the entropy, Eq.~\eqref{eq:entropy}, is maximized 
subject to some problem-specific constraints. 
In our case, they are i) the PSD estimate has to be non-negative; ii) its Fourier transform has to match the sample autocorrelation (wherever an estimate of this is available).

Before doing so, there is a technical problem to solve: the definition of entropy (Eq.~\ref{eq:entropy}) depends on a probability distribution, 
not on the PSD.
It can be proved  \cite{AblesMESA, Bartlett} that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of filtering a white noise process with a filter with transfer function $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising eq.~\eqref{eq:EntropyGain} is equivalent to maximising eq.~\eqref{eq:entropy_continuos}.

@@

Eq.~\eqref{eq:ARspectrum} and Eq.~\eqref{eq:MESApsd} are equivalent, if we identify $b_i = - a_i$ and $P_N \Delta t= \sigma ^ 2$.
This shows that the MAXENT estimation of the PSD also computes the coefficients of the autoregressive 
process that model the observed process.

\subsection{Forecasting} \label{sec:forecasting}
The link between MESA and autoregressive processes is of particular interest. Given the solution to the Levinson recursion to determine the $a_k$, see Appendix A, we have automatically the coefficients of the AR with equivalent PSD, hence we are able to exploit Eq.~\ref{eq:AR_p} to perform \emph{forecasting}, thus providing plausible future observations, conditioned on the observed data.
Indeed, an AR($p$) process models the conditional probability $p(x_t|x_{t-1}, \ldots , x_{t-p})$ of the observation at time $t$ with respect to the past $p$ observation:
\begin{align}\label{eq:p_forecast}
	p&(x_t|x_{t-1}, \ldots , x_{t-p}) \nonumber\\
	&= \frac{1}{\sigma\sqrt{2\pi}} \exp\left[-\frac{1}{2} \left(\frac{x_t - \sum_{i = 1}^p b_i x_{t-i}}{\sigma}\right)^2\right].
\end{align}
The interpretation of Eq.~(\ref{eq:p_forecast}) is straightforward: $x_t$ follows a Gaussian distribution with a fixed variance an a mean value $m_t = \sum_{i = 1}^p b_i x_{t-i}$ computed from past observations.
Eq.~(\ref{eq:p_forecast}) provides then a well defined probability framework for predicting future observations: this is a very useful feature of MESA, that does not have an equivalent in any other spectral analysis computation.

\section{Validation of the model}\label{sec:validation}
It is now clear that MESA provides a recursive formula for computing the coefficients $a_k$ in Eq.~(\ref{eq:MESApsd}). The number $M$ of such coefficients is the same as the number of time steps considered in the computation of the autocorrelation $\bar{r}_m$. In a perfect scenario, this would be equal to the number of points the autocorrelation is computed at (equivalent to the length of data considered). However, the computation of high order coefficients of the autocorrelation is unstable and for high enough $n$, the the estimation for  $\bar{n}_m$ shows a very high variance, broadly scaling as $\sim \left(\sqrt{M - m}\right)^{-1}$.

@@

It is then clear that the choice of how many samples of the discrete autocorrelation to consider is important: 
on the one hand it is advisable to include as much knowledge of the autocorrelation as possible, leading to include all the known $\bar{r}_m$; on the other hand, including values of the autocorrelation that are not reliably estimated, can be counterproductive.
Indeed, the order $M$ of the autocorrelation to be considered (or equivalently the order $M$ of the underlying autoregressive process) is the only tuning parameter of MESA and a careful balance between this two necessities must be made when applying the algorithm.

@@

Guided from numerical experiments, an indication on the upper bound to the autoregressive order $M_{max}$ is~\cite{doi:10.1190/1.1440902}:
\begin{equation}\label{eq:MMAx}
M_{max} = 2N / \ln{(2N)}\,,
\end{equation}
where $N$ is the number of points in the time-series.
However, this is just a plausible upper limit on the order of the AR process $m$ and the optimal algorithm could employ fewer points.
We then need a more sophisticated method for computing the right value for $m$.
Various loss functions to assess the algorithm performance have been proposed in literature.
We summarise them below:

\begin{itemize}
\item \textbf{Final prediction Error } 
The first criterion is due to Akaike~\cite{Akaike1970StatisticalPI}. He proposed that $m$ should be chosen as the 
length that minimises the error when the filter is used as a predictor, the \emph{final prediction} error (FPE): 
\begin{equation}
    FPE(m) = \mathbb{E}\left[ \left((x_t - \hat x_t) ^ 2\right) \right]
\end{equation}
with $\hat{x}_t = \sum_{i = 1}^M a_i x_{t - i}$.
Minimizing FPE is equivalent to minimizing the quantity: 
\begin{equation}
    \mathcal{L}_{\rm FPE}(m) = P_{m} \frac{N + m + 1}{N - m - 1}
\end{equation}
with $P_m$ defined by the prediction error filter equation \ref{eq:errorFilter1}. \\ 
If the $N \to \infty$ limit is considered,remembering $m_{max} = 2N / \log(2N)$, we see that N dominates both the numerator and the denominator of the fraction: in large $N$ limit using Akaike's loss function is equivalent to the minimization of the variance $P_m$ of the process.

@@

\item \textbf{Optimum Bayes Decision rule} 
The last criterion we will consider is the Optimum Bayes Decision rule (OBD)\cite{doi:10.1029/WR018i004p01097}. Let $x(t)$ be the observed 
time series for the process that has to be described as an autoregressive process of order $m$ (AR(m)) with m to be determined. The OBD is obtained chosing between M different hypothesis $H_i, i = 1, \dots, M$ maximizing their posterior distribution $P(H_i\vert x(t))$. Each
hypothesis is uniquely determined from both the length and the values of the filter under study. Chosing a Likelihood of the form \eqref{eq:p_forecast} and uniform priors for both the coefficients and the hypothesis Rao has shown that chosing the minimum for $-log(P(H_i \vert x(t))$, i.e. maximizing the posterior for $H_i$, is asymptotically equivalent to the minimization of:

@@

\subsection{Choice of the loss function: Gaussian PSD} \label{sec:arp_validation}
We test the performance of the three loss functions on a random time series generated with a known power spectral density.
In this first experiment, we take the PSD to be a Gaussian distribution with mean $\mu = 2.5 \quad units$ and standard deviation $\sigma = 0.5 \quad units$.
The time series are generated by sampling a frequency vector from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$, following the procedure outlined in sec.~\ref{sec:forecasting}, and performing an inverse Fast Fourier Transform on the frequency series. 
For this investigation, we generate a dataset of 1000 time series of 3000 points each.

We then apply the Burg method choosing in turn each of the loss functions on an ensemble of simulated time series, 
thus obtaining an ensemble of PSD estimates, that will allows the statistical characterisation of the performance of each loss function.
The disagreement between the PSD $S_i(f)$ estimated from the $i$-th simulated time series and the target PSD $S(f)$ is measured via 
the relative error $r_i$, averaged over all frequencies:
\begin{equation}\label{eq:freq_error}
r_i = \frac{1}{Ny}\sum_{f_j=0}^{f_j = Ny} \frac{|S_i(f_j) - S(f_j)|}{S(f_j)}
\end{equation}
where $Ny$ is the Nyquist frequency of the time series.

>>>>

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ABSTRACT
\begin{abstract}
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package `memspectrum`. We characterize our implementation by performing a power spectral density analysis on syntethic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using both syntethic data and real gravitational wave data from the LIGO-Virgo collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime.

@@

The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, astronomy as well as engineering and finance. Of particular interest for several areas of physics is the study of stochastic time series. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\cite{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will explore further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

@@

The choice of a window function is arbitrary and typically is made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of PSD is reached. A high frequency resolution implies high variance and vice-versa.
Besides the window function, Welch's method requires a number of arbitrary choices to be made, such as the number of time slices and the overlap between consecutive slices. All these knobs must be tuned by hand and their choice can dramatically affect the PSD estimation, hence begging the question of what the ``real" PSD is.

@@

In this paper, we discuss the details of the Maximum entropy principle, its application to the problem of PSD estimation with Burg algorithm and the link between Burg's algorithm and autoregressive process.
Our goal is to bring (again) to public attention Maximum Entropy Spectral analysis, in the hope that it will be widely employed as a way out to the many problems posed by the Welch's algorithm (or other similar methods).
To facilitate this goal, we present and describe a new code, \texttt{memspectrum}, that provides a robust and easy-to-use python implementation of the algorithm\footnote{
Available at link: \url{https://pypi.org/project/memspectrum/}.
}.
We provide a thorough assessment of the performance of our code and we validate our results performing a number of tests on simulated and real data.
We also compare our results with those of spectral analysis carried out with the standard Welch's method.
In order to apply our model on a realistic setting, we analyse some time series of broad interest in the scientific community.

Our paper is organized as follows: we being by briefly reviewing the theoretical foundations of the maximum entropy principle in Sec.~\ref{sec:foundations}. Sec.~\ref{sec:validation} presents the validation of Burg's method as well as of our implementation on simulated data. In Sec.~\ref{sec:Welch_comparison} we compare the results from \texttt{memspectrum} with the Welch method; Sec.~\ref{sec:applications} presents a few applications to time series and, finally, we conclude with a discussion in Sec.~\ref{sec:conclusion}.

@@

It is very common that the constraints provided by the evidence are not enough for setting the probabilities for each event: in this case, it is reasonable to assume that the probability assignment should make the experiment as unpredictable as possible\footnote{
In ref.~\cite{Jaynes_MAXENT} this statement is made more precise and justified more thoroughly, with arguments based on combinatorial analysis.
}.
In other words, the amount of `information' introduced by the the probability assignment should be as high as possible.

MAXENT puts the aforementioned principle into a more precise mathematical context by stating that probabilities should be assigned by maximizing uncertainty (information entropy) using evidence as constraint. 
This defines a variational problem, where the information entropy $H\left[p_1, \dots, p_N\right]$, defined in eq.~\eqref{eq:entropy}, has to maximized.

@@

Of particular importance is the case in which, in addition to the mean, also the variance is known. MAXENT leads to the Gaussian distribution. 
We retain this particularly interesting from the foundational point of view, since it provides a deeper insight into the ubiquitous Gaussian distribution.
Indeed, the Gaussian distribution is not only the limit distribution provided by the central limit theorem for finite variance processes but it is also the distribution that maximizes the entropy. For this reason, appealing to MAXENT principle, it is the correct assignment if mean and covariance are the only quantities that fully define our process. In some sense, we can interpret the central limit theorem as the natural `statistical' evolution toward a configuration that maximizes entropy.

For this work, we are particularly interested in the mutli-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realisation of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of autocovariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution\cite{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

For a wide-sense stationary process the mean function is independent of time, hence it can be redefined to be equal to zero without loss of generality, and the auto-covariance function is dependent only on the time lag $\tau \equiv t_i - t_j$. One can thus choose a sampling rate $\Delta t$ so that $C_{ij} = C((i-j)\Delta t)$. The autocovariance matrix thus becomes a Toeplitz matrix\footnote{
We remind the reader that a Toeplix matrix is a matrix in the form:
\begin{pmatrix}
	a_0 & a_1 & a_2 & \ldots & \ldots& \ldots &a_n\\
	a_{-1} & a_0 & a_1 & \ldots & \ldots& \ldots &a_{n-1}\\
	a_{-2} & a_{-1} & a_0 & \ldots & \ldots & \ldots  &a_{n-2}\\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots &\vdots\\
	a_{-n +1} & \ldots & \ldots & \ldots& a_{-1} & a_0    &a_{1}\\
	a_{-n} & \ldots & \ldots & \ldots& a_{-2} & a_{-1} & a_0
\end{pmatrix}

}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalised by the discrete Fourier transform base\cite{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} \tilde{x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\cite{prob_information_theory}
%P. M. Woodward, D. W. Fry, and W. Higinbotham, Probability and information theory, with applications to radar; 2nd ed. (Pergamon Press, Oxford, 1964), URL http://cds.cern.ch/record/2031792.
and of gravitational waves data analysis, e.g. \cite{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the autocovariance function.

@@

As in the previous examples, one needs to set up a variational problem where the entropy eq.~\eqref{eq:entropy} is maximized 
subject to some problem-specific constraints. 
In our case, they are i) the PSD estimate has to be non-negative; ii) its Fourier transform has to match the sample autocorrelation (wherever an estimate of this is available).

Before doing so, there is a technical problem to solve: the definition of entropy depends on a probability distribution, not on the PSD.
It can be proved \cite{AblesMESA, Bartlett} that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of filtering a white noise process with a filter with transfer function $T(f)$ equal to $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising eq.~\eqref{eq:EntropyGain} is equivalent to maximising eq.~\eqref{eq:entropy_continuos}.

@@

Eq.~\eqref{eq:ARspectrum} and eq.~\eqref{eq:MESApsd} are equivalent, if we identify $b_i = - a_i$ and $P_N \Delta t= \sigma ^ 2$.
This shows that the MAXENT estimation of the PSD also computes the coefficients of the autoregressive 
process that model the observed process.

\subsection{Forecasting} \label{sec:forecasting}
The link between MESA and autoregressive processes is of particular interest. Given the solution to the Levinson recursion to determine the $a_k$, see Appendix~\ref{sec:appendix}, we authomatically know the coefficients of the equivalent AR process, hence we are able to exploit eq.~\ref{eq:AR_p} to perform \emph{forecasting}, thus providing plausible future observations, conditioned on the observed data.
Indeed, for an AR($p$) process the conditional probability $p(x_t|x_{t-1}, \ldots , x_{t-p})$ of the observation at time $t$ with respect to the past $p$ observation has the form:
\begin{align}\label{eq:p_forecast}
	p&(x_t|x_{t-1}, \ldots , x_{t-p}) \nonumber\\
	&= \frac{1}{\sigma\sqrt{2\pi}} \exp\left[-\frac{1}{2} \left(\frac{x_t - \sum_{i = 1}^p b_i x_{t-i}}{\sigma}\right)^2\right].
\end{align}
The interpretation of Eq.~(\ref{eq:p_forecast}) is straightforward: $x_t$ follows a Gaussian distribution with a fixed variance an a mean value $m_t = \sum_{i = 1}^p b_i x_{t-i}$ computed from past observations.
Eq.~(\ref{eq:p_forecast}) provides then a well defined probability framework for predicting future observations: this is a very useful feature of MESA, that does not have an equivalent in any other spectral analysis computation.

\section{Validation of the model}\label{sec:validation}
It is now clear that MESA provides a recursive formula for computing the coefficients $a_k$ in Eq.~(\ref{eq:MESApsd}). The number $M$ of such coefficients is the same as the number of time steps considered in the computation of the autocorrelation $\bar{r}_m$. In a perfect scenario, this would be equal to the number of points the autocorrelation is computed at (equivalent to the length of data considered). However, the computation of high order coefficients of the autocorrelation is unstable and for high enough $m$, as the estimation for  $\bar{r}_m$ shows a very high variance, broadly scaling as $\sim \left(\sqrt{M - m}\right)^{-1}$.

@@

It is then clear that the choice of how many samples of the discrete autocorrelation to consider is important: 
on the one hand it is advisable to include as much knowledge of the autocorrelation as possible, leading to include all the known $\bar{r}_m$; on the other hand, including values of the autocorrelation that are not reliably estimated, can be counterproductive.
The order $M$ of the autocorrelation to be considered (or equivalently the order $M$ of the underlying autoregressive process) is the only tuning parameter of MESA and a careful balance between this two necessities must be made when applying the algorithm.

@@

Guided from numerical experiments, an indication on the upper bound to the autoregressive order $M_{max}$ is~\cite{doi:10.1190/1.1440902}:
\begin{equation}\label{eq:MMAx}
M_{max} = 2N / \ln{(2N)}\,,
\end{equation}
where $N$ is the number of observed points in the time-series.
However, this is just a plausible upper limit on the order of the AR process $m$ and the optimal algorithm could employ fewer points.
We then need a more sophisticated method for computing the right value for $m$.
Various loss functions to assess the algorithm performance have been proposed in literature.
We summarise them below:

\begin{itemize}
\item \textbf{Final prediction Error } 
The first criterion is due to Akaike~\cite{Akaike1970StatisticalPI}. He proposed that $m$ should be chosen as the 
length that minimises the error when the filter is used as a predictor, the \emph{final prediction} error (FPE): 
\begin{equation}
    FPE(m) = \mathbb{E}\left[ \left((x_t - \hat x_t) ^ 2\right) \right]
\end{equation}
with $\hat{x}_t = \sum_{i = 1}^M a_i x_{t - i}$.
Minimizing FPE is equivalent to minimizing the quantity: 
\begin{equation}
    \mathcal{L}_{\rm FPE}(m) = P_{m} \frac{N + m + 1}{N - m - 1}
\end{equation}
with $P_m$ being the estimated noise variance at order $m$, see eq.~\eqref{eq:errorFilter1}.

In the $N \to \infty$ limit, remembering $m_{max} \sim 2N / \log(2N)$, Akaike's loss function is equivalent to the minimization of the variance $P_m$ of the white noise of the underlying $AR(p)$ model.

@@

\item \textbf{Optimum Bayes Decision rule} 
The last criterion we will consider is the Optimum Bayes Decision rule (OBD)\cite{doi:10.1029/WR018i004p01097}. Let $x(t)$ be the observed 
time series for the process that has to be described as an autoregressive process of order $m$ ($AR(m)$), with $m$ to be determined. The OBD is obtained chosing between M different hypothesis $H_i, i = 1, \dots, M$ maximizing their posterior distribution $P(H_i\vert x(t))$. Each
hypothesis is uniquely determined from both the length and the values of the filter under study. Chosing a Likelihood of the form \eqref{eq:p_forecast} and uniform priors for both the coefficients and the hypothesis Rao has shown that chosing the minimum for $-log(P(H_i \vert x(t))$, i.e. maximizing the posterior for $H_i$, is asymptotically equivalent to the minimization of:

@@

\subsection{Choice of the loss function: Gaussian PSD} \label{sec:arp_validation}
We test the performance of the three loss functions on a random time series generated with a known power spectral density.
In this first experiment, we take the PSD to be a Gaussian distribution with mean $\mu = 2.5 \quad units$ and standard deviation $\sigma = 0.5 \quad units$.
The time series are generated by sampling a frequency vector from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$ and performing an inverse Fast Fourier Transform on the frequency series. 
For this investigation, we generate a dataset of $N=1000$ time series of $3000$ points each.

We then apply the Burg method choosing in turn each of the loss functions on an ensemble of simulated time series, 
thus obtaining an ensemble of PSD estimates, that will allows the statistical characterisation of the performance of each loss function.
The disagreement between the PSD $S_i(f)$ estimated from the $i$-th simulated time series and the target PSD $S(f)$ is measured via 
the \textit{frequency-averaged relative error} $r_i$:
\begin{equation}\label{eq:freq_error}
	r_i = \frac{1}{N_f}\sum_{f_j=\{0, \hdots , Ny\}} \frac{|S_i(f_j) - S(f_j)|}{S(f_j)}
\end{equation}
where $Ny$ is the Nyquist frequency of the time series and $N_f$ is the number of the discrete frequencies the PSD is evaluated at.

For each loss function, we compute the ensemble-averaged PSD (as well as the $90\%$ confidence level) and the \textit{ensemble-averaged relative error}:
\begin{equation}\label{eq:ensemble_averaged_error}
	r(f) = \frac{1}{N} \sum_i \frac{|S_i(f) - S(f)|}{S(f)} \;.
\end{equation}
For each loss function, figs.~\ref{fig:FPEmean} \ref{fig:OBDmean} and \ref{fig:CATmean} show the averaged PSD, 
as well as the ensemble-averaged relative error from Eq.~\eqref{eq:ensemble_averaged_error}.
Furthermore, Fig.~\ref{fig:optcomparison} displays the relation between $r_i$ and the autoregressive 
order $m_i$ chosen for each independently drawn time series.
